Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-24 bbbq_article.tex

To determine if epitopes derived from TMHs are presented 
we used two elution studies, which both have sequenced
epitopes found in naturally occurring human immune cells,
for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different}).
For each of the sequenced epitopes,
we found its possible location(s) in 
a human representative reference proteome,
with UniProt ID UP000005640\_9606.
For the epitopes that were found exactly once in the proteome,
we predicted the topology of the proteins in which the epitopes are found.
We used both TMHMM and PureseqTM to do these predictions. 
From this topology, we could determine if the epitope
overlapped with a transmembrane regions.

@@

For MHC-I, 9-mers were used, as this is the length used in our earlier
study (\cite{bianchi2017}). For MHC-II, 14-mers were used, 
as these are the most frequently occurring
epitope length (\cite{bergseng2015different}).
We used a human (UniProt ID UP000005640\_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (Mycobacterium tuberculosis (MTb), UniProt ID UP000001584) 
reference proteome. The pathogens were chosen based 
on their impact on the human 
population \richel{ref}.
We used TMHMM  \richel{ref} to predict the topology 
of the proteins within the proteome.
To predict the affinity of an epitope to a certain MHC molecule,
we used \verb;EpitopePrediction; (\cite{bianchi2017}) for MHC-I 
and \verb;MHCnuggets; \richel{ref} for MHC-II.
The 13 MHC-I haplotypes used in this study are the same as 
used in our previous study (\cite{bianchi2017}).
The MHC-II haplotypes used additionally were selected 
to occur with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional}),
resulting in 21 haplotypes.

@@

To detect the evolutionary conservation of TMHs,
we collected human single nucleotide polymorphisms (SNPs)
that resulted in a single amino the substitution,
and tallied if this substitution occurred within a TMH or not.
From that tally, we determined how likely the number of SNPs
found in TMHs is by chance.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used.
Pivotal was the 'dbSNP' (\cite{sherry2001dbsnp}) database,
which contains 650 million 
catalogued non-redundant humane variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}).
Additionally, the aptly-named 'gene' (for gene names)
and 'protein' (for proteins sequences) database were used.

@@

The number of variations
was limited to the first 250 variations per gene (see figure
\ref{fig:snps_per_gene_name_ncbi} for the number of SNPs
within the dbSNP database for these gene names),
resulting in 60,683 variations at the protein level.
Only variations that result in a SNP for
a single amino acid substitution were analyzed, resulting in 38,882 SNPs.

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that ID number is unrelated to SNP positions the relative positions of the analyzed SNPS in a protein was determined and showed no positional bias, shown in figure \ref{fig:snp_rel_pos}.

>>>>

To determine if epitopes derived from TMHs are presented 
we used two elution studies, which both have sequenced
epitopes found in naturally occurring human immune cells,
for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different}).
For each of the sequenced epitopes,
we found its possible location(s) in 
a human representative reference proteome,
with UniProt ID UP000005640_9606.
For the epitopes that were found exactly once in the proteome,
we predicted the topology of the proteins in which the epitopes are found.
We used both TMHMM and PureseqTM to do these predictions. 
From this topology, we could determine if the epitope
overlapped with a transmembrane regions.

@@

For MHC-I, 9-mers were used, as this is the length used in our earlier
study (\cite{bianchi2017}). For MHC-II, 14-mers were used, 
as these are the most frequently occurring
epitope length (\cite{bergseng2015different}).
We used a human (UniProt ID UP000005640_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (Mycobacterium tuberculosis (MTb), UniProt ID UP000001584) 
reference proteome. The pathogens were chosen based 
on their impact on the human 
population \frans{Add references here}.
We used TMHMM  \richel{ref} to predict the topology 
of the proteins within the proteome.
To predict the affinity of an epitope to a certain MHC molecule,
we used \verb;EpitopePrediction; (\cite{bianchi2017}) for MHC-I 
and \verb;MHCnuggets; \richel{ref} for MHC-II.
The 13 MHC-I haplotypes used in this study are the same as 
used in our previous study (\cite{bianchi2017}).
The MHC-II haplotypes used additionally were selected 
to occur with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional}),
resulting in 21 haplotypes.

@@

To detect the evolutionary conservation of TMHs, 
single nucleotide
polymorphisms (SNPs) within the human population
that resulted in the substitution of one amino acid were collected from ??? and tallied if
the SNP occurred within a TMH or not. Using statistics is was determined if there are more, less or an equal amount of SNPs in TMHs
as a measure of evolutionary conservation.
This workflow is discussed in more detail below.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used,
which are 'gene',\frans{snap ik niet which are gene?} to find the gene names of membrane proteins, 
'dbSNP' (\cite{sherry2001dbsnp}) for SNPs associated with those genes
and 'protein', to obtain the sequence of proteins that SNPs act upon.
The 'dbSNP' contains 650 million 
catalogued non-redundant humane variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}).

% \paragraph{Data retrieval}

\frans{ dit is relevant waarom: ? To retrieve the data from these databases the
\verb;rentrez; R package (\cite{rentrez}) was used
that calls the NCBI website's API. To provide for a 
stable user experience for all users, 
this API limits the user to 3 calls per second.
Additionally, the API splits the result of a bigger
query into multiple pages, each of which needs one API call.
We wrote the \verb;sprentrez; package (\cite{sprentrez}) to provide for 
bigger queries of multiple (and delayed) API calls.}

@@

The number of variations
was limited to the first 250 variations per gene (see figure
\ref{fig:snps_per_gene_name_ncbi} for the number of SNPs
within the dbSNP database for these gene names),
resulting in 60,683 variations at the protein level.
Only variations that result in a SNP for
a single amino acid substitution were analysed, resulting in 38,882 SNPs.

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that ID number is unrelated to SNP positions the relative positons of the analyzed SNPS in a protein was determined and showed no positional bias, shown in figure \ref{fig:snp_rel_pos}.
